---
title: "SDIS Summary Analysis"
author: "Tina Lasisi | Edited: João P. Donadio"
date: today
format:
  html:
    code-fold: false
    toc: true
execute:
  echo: true
  warning: false
---

## Overview

This analysis examines State DNA Index System (SDIS) data that includes information reported separately for each state's DNA database. The data captures key dimensions including:

- Total size of each state's DNA database
- Whether states collect DNA from arrestees (not just convicted offenders)
- Whether states allow familial DNA searching
- References to relevant state statutes (from Murphy & Tong appendix)

This information provides insight into the variation in DNA database policies, practices, and legal frameworks across U.S. states.

## Setup

Load necessary packages for the analysis.

```{r}
#| label: packages-setup
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show setup code"

# List of required packages
required_packages <- c(
  "tidyverse",    # Data manipulation and visualization
  "readr",      # Reading CSV files
  "dplyr",      # Data manipulation
  "tidyr",      # Data tidying
  "purrr",      # Functional programming tools
  "ggplot2",    # Data visualization
  "heatmaply",  # Interactive heatmaps
  "kableExtra", # Enhanced tables for reporting
  "DT",         # Interactive tables
  "flextable",  # Flexible tables for reporting
  "maps",       # Mapping tools
  "here",       # File path management
  "remotes",    # To install urbnmapr package
  "sf",
  "patchwork"
)

# Function to install missing packages
install_missing <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      message(paste("Installing missing package:", pkg))
      install.packages(pkg, dependencies = TRUE)
    }
  }
}

# Install any missing packages
install_missing(required_packages)

# Load all packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(heatmaply)
  library(kableExtra)
  library(DT)
  library(flextable)
  library(maps)
  library(here)
  library(remotes)
  library(sf)
  library(patchwork)
})

# Verify all packages loaded successfully
loaded_packages <- sapply(required_packages, require, character.only = TRUE)
if (all(loaded_packages)) {
  message("📚 All packages loaded successfully!")
} else {
  warning("The following packages failed to load: ", 
          paste(names(loaded_packages)[!loaded_packages], collapse = ", "))
}

```

## Import SDIS data

Load the SDIS dataset and display a glimpse of its structure, including data types, missing values, and unique counts.

**Database Columns Definition**

| Column | Description |
|--------|-------------|
| state | Name of the U.S. state or jurisdiction |
| n_total | Total number of DNA profiles in the state database |
| n_arrestees | Number of arrestee DNA profiles in the database |
| n_offenders | Number of convicted offender DNA profiles in the database |
| n_forensic | Number of forensic/crime scene DNA profiles in the database |
| arrestee_collection | Whether the state collects DNA from arrestees ('yes'/'no') |
| collection_statute | Citation to the state's DNA collection statute |
| fam_search | Family search policy status ('permitted', 'prohibited', 'unspecified') |
| database_source | URL source for the database statistics |
| database_source_year | Year the database statistics were reported |
| verification_comment | Additional notes or comments about data verification |

```{r}
#| label: import-setup
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show import code"
#| message: false
#| warning: false

# Set up path to data file
data_file <- file.path(here("data", "sdis", "raw", "sdis_raw.csv"))

# Load the SDIS data
sdis_data <- read_csv(data_file)

# Display data types for each column
enhanced_glimpse <- function(df) {
  glimpse_data <- data.frame(
    Column = names(df),
    Type = sapply(df, function(x) paste(class(x), collapse = ", ")),
    Rows = nrow(df),
    Missing = sapply(df, function(x) sum(is.na(x))),
    Unique = sapply(df, function(x) length(unique(x))),
    First_Values = sapply(df, function(x) {
      if(is.numeric(x)) {
        paste(round(head(x, 3), 2), collapse = ", ")
      } else {
        paste(encodeString(head(as.character(x), 3)), collapse = ", ")
      }
    })
  )
  
  ft <- flextable(glimpse_data) %>%
    theme_zebra() %>%
    set_caption(paste("Enhanced Data Glimpse:", deparse(substitute(df)))) %>%
    autofit() %>%
    align(align = "left", part = "all") %>%
    colformat_num(j = c("Rows", "Missing", "Unique"), big.mark = "") %>%
    bg(j = "Missing", bg = function(x) ifelse(x > 0, "#FFF3CD", "transparent")) %>%
    bg(j = "Unique", bg = function(x) ifelse(x == 1, "#FFF3CD", "transparent")) %>%
    add_footer_lines(paste("Data frame dimensions:", nrow(df), "rows ×", ncol(df), "columns")) %>%
    fontsize(size = 10, part = "all") %>%
    set_table_properties(layout = "autofit", width = 1)
  
  return(ft)
}

enhanced_glimpse(sdis_data)

```

## Arrestee Collection Information

In the United States, the collection of DNA from individuals upon arrest, prior to any conviction, is a significant law enforcement practice with complex legal and ethical dimensions.

The following map and table illustrate the current patchwork of state laws governing this practice.

```{r}
#| label: arrestee-collection-visualization
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show arrestee collection visualization code"
#| fig-width: 12
#| fig-height: 8

# Create summary data for arrestee collection
arrestee_summary <- sdis_data %>%
  group_by(arrestee_collection) %>%
  summarise(
    count = n(),
    states = list(state),
    .groups = "drop"
  ) %>%
  arrange(desc(count))

# Get US state map data
us_states <- map_data("state")

# Prepare map data
map_data <- sdis_data %>%
  mutate(region = tolower(state)) %>%
  right_join(us_states, by = "region") %>%
  filter(!is.na(arrestee_collection))

# Option 1: US Map Visualization
map_plot <- ggplot(map_data, aes(x = long, y = lat, group = group, fill = arrestee_collection)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_fixed(1.3) +
  scale_fill_manual(
    values = c(
      "yes" = "#0072B2",      # Blue for collecting states
      "no" = "#666666"       # Gray for non-collecting states
    ),
    name = "Collects DNA from Arrestees",
    labels = c("No", "Yes")
  ) +
  labs(
    title = "DNA Collection from Arrestees by State",
    subtitle = "States that collect DNA samples from individuals upon arrest"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40", size = 12),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

# Create detailed summary table
summary_table <- arrestee_summary %>%
  mutate(
    state_list = map_chr(states, ~paste(.x, collapse = ", ")),
    percentage = round(count / sum(count) * 100, 1)
  ) %>%
  select(Status = arrestee_collection, Count = count, Percentage = percentage, States = state_list)

# Display the visualizations
print(map_plot)

# Create a nice flextable for the report
ft <- flextable(summary_table) %>%
  theme_zebra() %>%
  set_caption("Arrestee DNA Collection Status by State") %>%
  autofit() %>%
  align(align = "left", part = "all") %>%
  bg(j = "Status", bg = function(x) {
    case_when(
      is.na(x) ~ "#CCCCCC",
      x == "yes" ~ "#1f77b4",
      x == "no" ~ "#666666",
      TRUE ~ "#CCCCCC"
    )
  }) %>%
  color(j = "Status", color = "white") %>%
  bold(j = "Status") %>%
  fontsize(size = 11, part = "all")

ft
```

### Adjusting `n_arrestees`

- Adjust the dataset to ensure consistency in arrestee DNA collection reporting.

- Setting `n_arrestees` to 0 for states that do not collect arrestee DNA, even if they report non-zero values.

- Removed the `arrestee_colection` column after adjustments were made.

```{r}
#| label: preprocessing-data
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show pre-processing code"
# Create a copy of the data for processing
sdis_data_processed <- sdis_data

# Count states affected by this adjustment
states_with_no_collection <- sdis_data_processed %>% 
  filter(arrestee_collection == 'no')

states_to_adjust <- states_with_no_collection %>% 
  filter(!is.na(n_arrestees) & n_arrestees != 0)

if (nrow(states_to_adjust) > 0) {
  cat("States with arrestee_collection='no' but non-zero n_arrestees values:\n")
  for (i in 1:nrow(states_to_adjust)) {
    state <- states_to_adjust[i, ]
    cat(paste0("  • ", state$state, ": n_arrestees = ", format(state$n_arrestees, big.mark = ","), "\n"))
  }
}

# Set n_arrestees to 0 for states that don't collect arrestee DNA
sdis_data_processed <- sdis_data_processed %>%
  mutate(n_arrestees = ifelse(arrestee_collection == 'no', 0, n_arrestees))

# Use processed data for all subsequent analyses
sdis_data <- sdis_data_processed
```

## Family Search Policy Status

Familial DNA searching is an advanced forensic technique that uses partial DNA matches to identify potential relatives of an unknown suspect in a criminal database.

The following visualization maps the current patchwork of state policies ('permitted', 'prohibited', 'unspecified').

```{r}
#| label: family-search-visualization
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show family search policy visualization code"
#| fig-width: 12
#| fig-height: 10

# Create summary data for family search policies
family_search_summary <- sdis_data %>%
  mutate(
    fam_search = factor(fam_search, 
                       levels = c("permitted", "prohibited", "unspecified"),
                       labels = c("Permitted", "Prohibited", "Unspecified")),
    fam_search_simple = case_when(
      fam_search == "Permitted" ~ "Permitted",
      fam_search == "Prohibited" ~ "Prohibited",
      TRUE ~ "Unspecified/No Policy"
    )
  ) %>%
  group_by(fam_search, fam_search_simple) %>%
  summarise(
    count = n(),
    states = list(state),
    .groups = "drop"
  ) %>%
  arrange(desc(count))

# Get complete data for mapping (include states with missing data)
all_states_map <- map_data("state") %>%
  as_tibble() %>%
  distinct(region) %>%
  mutate(
    state_name = str_to_title(region),
    has_data = state_name %in% sdis_data$state
  )

# Prepare map data for family search policies
family_search_map_data <- sdis_data %>%
  mutate(
    region = tolower(state)) %>%
  right_join(map_data("state"), by = c("region" = "region")) %>%
  filter(!is.na(fam_search))

# Option 1: US Map Visualization
family_map_plot <- ggplot(family_search_map_data, 
                         aes(x = long, y = lat, group = group, fill = fam_search)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_fixed(1.3) +
  scale_fill_manual(
    values = c(
      "permitted" = "#1f77b4",    # Blue for permitted
      "prohibited" = "#000000ff",   # Black for prohibited
      "unspecified" = "#666666"   # Gray for unspecified
    ),
    name = "Familial Search Policy",
    drop = FALSE
  ) +
  labs(
    title = "Familial DNA Search Policies by State",
    subtitle = "State policies regarding familial DNA searching in databases",
    caption = "Source: SDIS Database Analysis"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40", size = 12),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

# Display the visualizations
print(family_map_plot)

# Create a comprehensive summary table
summary_table_data <- family_search_summary %>%
  mutate(
    state_list = map_chr(states, ~paste(.x, collapse = ", ")),
    percentage = round(count / sum(count) * 100, 1)
  ) %>%
  select(Policy = fam_search, Count = count, Percentage = percentage, States = state_list)

# Create a nice flextable for the report
ft <- flextable(summary_table_data) %>%
  theme_zebra() %>%
  set_caption("Familial DNA Search Policy Status by State") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "States") %>%
  bg(j = "Policy", bg = function(x) {
    case_when(
      x == "Permitted" ~ "#1f77b4",
      x == "Prohibited" ~ "#000000ff",
      TRUE ~ "#666666"
    )
  }) %>%
  color(j = "Policy", color = "white") %>%
  bold(j = "Policy") %>%
  fontsize(size = 11, part = "all") %>%
  width(j = "States", width = 3)

ft
```

## Three-Panel Policy Map: Horizontal Layout

This visualization presents three key policy dimensions side-by-side: arrestee DNA collection, familial search policies, and FOIA data availability.

```{r}
#| label: three-panel-map
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show three-panel map code"
#| fig-width: 18
#| fig-height: 6

remotes::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)

# Create FOIA availability column
sdis_data <- sdis_data %>%
  mutate(
    foia_availability = case_when(
      state %in% c("California", "Florida", "Indiana", "Maine", 
                   "Nevada", "South Dakota", "Texas") ~ "provided",
      TRUE ~ "not_provided"
    )
  )

# Load US map data with proper AK/HI positioning
us_map_data <- get_urbn_map("states", sf = TRUE)

# Prepare map data
sdis_data_map <- sdis_data %>%
  mutate(region = tolower(state))

us_map_data <- us_map_data %>%
  mutate(region = tolower(state_name))

# Merge polygon data with policy data
map_data_repos <- us_map_data %>%
  left_join(sdis_data_map, by = "region")

# Function to create individual maps
create_policy_map <- function(fill_var, fill_title, fill_colors, plot_title) {
  ggplot(map_data_repos) +
    geom_sf(aes(fill = !!sym(fill_var)), color = "white", linewidth = 0.3) +
    scale_fill_manual(
      name = fill_title,
      values = fill_colors,
      na.value = "gray90",
      na.translate = FALSE,
      guide = guide_legend(
        direction = "vertical",
        title.position = "top",
        title.hjust = 0.5
      )
    ) +
    theme_void() +
    labs(title = plot_title) +
    theme(
      plot.title = element_text(size = 25, face = "bold", hjust = 0.5, margin = margin(b = 5)),
      legend.position = "bottom",
      legend.title = element_text(size = 1, face = "bold"),
      legend.text = element_text(size = 25),
      legend.box = "vertical",
      legend.margin = margin(t = 5, b = 0)
    )
}

# Create the three maps
map_arrestee_repos <- create_policy_map(
  "arrestee_collection", " ",
  c("yes" = "#1f77b4", "no" = "#666666"), 
  "A. Arrestee DNA Collection Policy"
)

map_familial_repos <- create_policy_map(
  "fam_search", " ",
  c("permitted" = "#1f77b4", "prohibited" = "#000000ff", "unspecified" = "#666666"),
  "B. Familial Search Policy"
)

map_foia_repos <- create_policy_map(
  "foia_availability", " ",
  c("provided" = "#1f77b4", "not_provided" = "#666666"),
  "C. FOIA Response Status"
)

# Combine maps horizontally
figure_combined <- map_arrestee_repos | map_familial_repos | map_foia_repos

figure_combined + plot_annotation(
  title = " ",
  theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
)

# Compute summary counts and percentages for each category
policy_summary <- sdis_data %>%
  summarise(
    `Arrestee Collection - Yes` = sum(arrestee_collection == "yes", na.rm = TRUE),
    `Arrestee Collection - No` = sum(arrestee_collection == "no", na.rm = TRUE),
    `Familial Search - Permitted` = sum(fam_search == "permitted", na.rm = TRUE),
    `Familial Search - Prohibited` = sum(fam_search == "prohibited", na.rm = TRUE),
    `Familial Search - Unspecified` = sum(fam_search == "unspecified", na.rm = TRUE),
    `FOIA - Provided` = sum(foia_availability == "provided", na.rm = TRUE),
    `FOIA - Not Provided` = sum(foia_availability == "not_provided", na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Count") %>%
  mutate(
    Group = case_when(
      grepl("Arrestee Collection", Metric) ~ "Arrestee Collection",
      grepl("Familial Search", Metric) ~ "Familial Search",
      grepl("FOIA", Metric) ~ "FOIA Availability"
    ),
    Percentage = round(Count / 50 * 100, 1)
  )

# Split into subtables by group
policy_tables <- policy_summary %>%
  group_split(Group)

# Create function to build consistent flextables
make_policy_table <- function(df) {
  flextable(df %>% select(Metric, Count, Percentage)) %>%
    theme_zebra() %>%
    set_caption(paste0(unique(df$Group), " (n = 50 states)")) %>%
    colformat_num(j = "Percentage", suffix = "%") %>%
    align(align = "left", part = "all") %>%
    bold(j = "Metric") %>%
    fontsize(size = 10, part = "all") %>%
    autofit()
}

# Build three subtables
ft_1 <- make_policy_table(policy_tables[[1]])
ft_2 <- make_policy_table(policy_tables[[2]])
ft_3 <- make_policy_table(policy_tables[[3]])

ft_1

ft_2

ft_3
```

## Data Availability Overview

Assess the completeness of data across states, including which states are missing and the availability of key fields.

```{r}
#| label: overview-data
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show overview code"
#| fig-width: 10
#| fig-height: 30
# Identify states present in the dataset
states_in_data <- unique(sdis_data$state) %>% sort()

# Check if all 50 states are represented
all_states <- c('Alabama', 'Alaska', 'Arizona', 'Arkansas', 'California', 'Colorado', 'Connecticut',
                'Delaware', 'Florida', 'Georgia', 'Hawaii', 'Idaho', 'Illinois', 'Indiana', 'Iowa',
                'Kansas', 'Kentucky', 'Louisiana', 'Maine', 'Maryland', 'Massachusetts', 'Michigan',
                'Minnesota', 'Mississippi', 'Missouri', 'Montana', 'Nebraska', 'Nevada', 'New Hampshire',
                'New Jersey', 'New Mexico', 'New York', 'North Carolina', 'North Dakota', 'Ohio',
                'Oklahoma', 'Oregon', 'Pennsylvania', 'Rhode Island', 'South Carolina', 'South Dakota',
                'Tennessee', 'Texas', 'Utah', 'Vermont', 'Virginia', 'Washington', 'West Virginia',
                'Wisconsin', 'Wyoming')

missing_states <- setdiff(all_states, states_in_data)
if (length(missing_states) > 0) {
  cat(paste("\nMissing states:", paste(missing_states, collapse = ", "), "\n"))
} else {
  cat("\nAll 50 states are represented in the dataset\n")
}

# Assess data completeness for each state
# Improved heatmap with enhanced visualization
data_availability <- sdis_data %>%
  filter(!is.na(state)) %>%  # Remove any rows with missing state names
  group_by(state) %>%
  summarise(across(everything(), ~sum(!is.na(.))), .groups = "drop") %>%
  as.data.frame()

# Set row names safely
state_names <- data_availability$state
data_availability <- data_availability %>% select(-state)
rownames(data_availability) <- state_names

# Generate visualization of data completeness
key_fields <- c('n_total', 'n_arrestees', 'n_offenders', 'n_forensic')

# Filter to include only key fields that exist in the data
available_key_fields <- intersect(key_fields, names(data_availability))

if (length(available_key_fields) > 0) {
  availability_subset <- data_availability[, available_key_fields, drop = FALSE]
  availability_binary <- as.data.frame(ifelse(availability_subset > 0, 1, 0))
  
  # Create heatmap data in long format
  heatmap_long <- availability_binary %>%
    rownames_to_column("state") %>%
    pivot_longer(cols = -state, names_to = "field", values_to = "available") %>%
    mutate(
      state = factor(state, levels = rownames(availability_binary)),
      state = factor(state, levels = sort(unique(state), decreasing = TRUE)),
      field = factor(field, levels = available_key_fields),
      # Create labels with different symbols
      label = case_when(
        available == 1 ~ "✓",
        available == 0 ~ "!",
        TRUE ~ ""
      ),
      # Color coding for different values
      fill_color = case_when(
        available == 1 ~ "Available",
        available == 0 ~ "Missing",
        TRUE ~ "Unknown"
      )
    )
  
  # Create the enhanced heatmap
ggplot(heatmap_long, aes(x = field, y = state, fill = fill_color)) +
  geom_tile(color = "white", linewidth = 0.8, width = 0.9, height = 0.9) +
  geom_text(aes(label = label, color = fill_color), 
            size = 4, fontface = "bold", vjust = 0.8) +
  scale_fill_manual(values = c(
    "Available" = "#2E86AB",
    "Missing" = "#FF6B6B",
    "Unknown" = "#f0f0f0"
  )) +
  scale_color_manual(values = c(
    "Available" = "white",
    "Missing" = "white",
    "Unknown" = "gray30"
  )) +
  labs(
    title = "Data Availability Heatmap by State",
    subtitle = "✓ = Data available | ! = Data missing (0 values)",
    x = NULL,
    y = NULL,
    fill = "Data Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x.top = element_text(angle = 315, hjust = 0.5, face = "bold"), 
    axis.text.y = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold", vjust = 0),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40", size = 10),
    panel.grid = element_blank(),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    axis.ticks.x = element_line(),
    axis.ticks.x.top = element_line()
  ) +
  guides(color = "none") +
  coord_fixed() +
  scale_x_discrete(position = "top")
  
} else {
  message("No key fields available for heatmap visualization")
}

# Data coverage summary
cat("\nData field coverage across states:\n")

key_fields <- c('n_total', 'n_arrestees', 'n_offenders', 'n_forensic', 'fam_search', 'collection_statute')

for (col in key_fields) {
  if (col %in% names(sdis_data)) {
    states_with_data <- sum(!is.na(sdis_data[[col]]))
    coverage_pct <- states_with_data / length(states_in_data) * 100
    cat(paste0(col, ": ", states_with_data, " states (", round(coverage_pct, 1), "%)\n"))
  }
}
```

## Total Profile Calculations Verification

This section examines states reporting n_total alongside component counts to determine whether totals represent:

1. Sum of all profile types including forensic (n_arrestees + n_offenders + n_forensic)

2. Sum of combined profiles only (n_arrestees + n_offenders)

```{r}
#| label: total_profiles
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show calculations code"
# Identify states with n_total and at least one component count
# First ensure all numeric columns are properly typed
states_with_totals <- sdis_data %>%
  mutate(
    n_arrestees = as.numeric(n_arrestees),
    n_offenders = as.numeric(n_offenders),
    n_forensic = as.numeric(n_forensic),
    n_total = as.numeric(n_total)
  ) %>%
  filter(!is.na(n_total)) %>%
  mutate(
    # Calculate sums
    sum_all = rowSums(cbind(n_arrestees, n_offenders, n_forensic), na.rm = TRUE),
    sum_arrestees_offenders = rowSums(cbind(n_arrestees, n_offenders), na.rm = TRUE),
    # Check if all components are available
    all_components_available = !is.na(n_arrestees) & !is.na(n_offenders) & !is.na(n_forensic),
    both_arrestees_offenders = !is.na(n_arrestees) & !is.na(n_offenders)
  )

# Check matches with tolerance
tolerance <- 10

states_with_totals <- states_with_totals %>%
  mutate(
    matches_combined_forensic = ifelse(
      all_components_available,
      abs(n_total - sum_all) <= tolerance,
      FALSE
    ),
    matches_combined = ifelse(
      both_arrestees_offenders,
      (abs(n_total - sum_arrestees_offenders) <= tolerance) & 
        (n_arrestees > 0),
      FALSE
    )
  )

# Filter to states with at least one component count
has_components <- states_with_totals %>%
  filter(!is.na(n_arrestees) | !is.na(n_offenders) | !is.na(n_forensic))

cat(paste("States with n_total and component data:", nrow(has_components), "\n"))
cat("\nTotal calculation patterns:\n")

# Categorize states
includes_all <- has_components %>% filter(matches_combined_forensic) %>% pull(state)
forensic_only <- has_components %>% filter(matches_combined & !matches_combined_forensic) %>% pull(state)
neither <- has_components %>% filter(!matches_combined_forensic & !matches_combined) %>% pull(state)

if (length(includes_all) > 0) {
  cat("\nn_total matches combined profiles with forensic (arrestees + offenders + forensic):\n")
  for (state in includes_all) {
    cat(paste("  •", state, "\n"))
  }
} else {
  cat("\nStates where n_total matches the sum of arrestees + offenders + forensic: None\n")
}

if (length(forensic_only) > 0) {
  cat("\nStates where n_total includes combined profiles only (arrestees + offenders):\n")
  for (state in forensic_only) {
    cat(paste("  •", state, "\n"))
  }
}

if (length(neither) > 0) {
  cat("\nStates where n_total does not match calculated sums:\n")
  for (state in neither) {
    state_data <- has_components %>% filter(state == !!state)
    cat(paste0("  • ", state, ":\n n_total=", format(state_data$n_total, big.mark = ","), 
               "\n Sum_all=", format(state_data$sum_all, big.mark = ","),
               "\n Sum_arrestees_offenders =", format(state_data$sum_arrestees_offenders, big.mark = ","), "\n\n"))
  }
}

# Display detailed breakdown
cat("\nDetailed breakdown:\n")
has_components %>%
  select(state, n_total, n_arrestees, n_offenders, n_forensic, matches_combined_forensic, matches_combined) %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), "", format(., big.mark = ",")))) %>%
  flextable()

```

## Analysis of Database Totals and Data Quality Issues

Examine states where N_total values reveal potential data quality issues or reporting inconsistencies.

```{r}
#| label: data_quality_analysis
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show data quality assessment code"

# Create enhanced data quality analysis
sdis_enhanced <- sdis_data %>%
  mutate(
    # Ensure all numeric columns are properly typed
    n_total = as.numeric(n_total),
    n_offenders = as.numeric(n_offenders),
    n_arrestees = as.numeric(n_arrestees),
    n_forensic = as.numeric(n_forensic)
  ) %>%
  rename(n_total_reported = n_total)

# Calculate different total relationships with small tolerance
tolerance <- 10

sdis_enhanced <- sdis_enhanced %>%
  mutate(
    total_equals_offenders = ifelse(
      !is.na(n_total_reported) & !is.na(n_offenders) & n_offenders > 0,
      abs(n_total_reported - n_offenders) <= tolerance,
      FALSE
    ),
    total_equals_off_arr = ifelse(
      !is.na(n_total_reported) & !is.na(n_offenders) & !is.na(n_arrestees) &
        n_offenders > 0 & n_arrestees > 0,
      abs(n_total_reported - (n_offenders + n_arrestees)) <= tolerance,
      FALSE
    ),
    total_equals_all = ifelse(
      !is.na(n_total_reported) & !is.na(n_offenders) & !is.na(n_arrestees) & 
        !is.na(n_forensic) & n_offenders > 0 & n_arrestees > 0 & n_forensic > 0,
      abs(n_total_reported - (n_offenders + n_arrestees + n_forensic)) <= tolerance,
      FALSE
    ),
    total_method = case_when(
      total_equals_all ~ "All components",
      total_equals_off_arr ~ "Offenders + Arrestees",
      total_equals_offenders ~ "Offenders only",
      TRUE ~ "Unknown"
    )
  )

# Create n_total_estimated based on the rules specified
sdis_enhanced <- sdis_enhanced %>%
  mutate(
    n_total_estimated = NA_real_,
    n_total_estimated_comment = ""
  )

# Rule 1: States where n_total == n_offenders + n_arrestees
mask_off_arr <- sdis_enhanced$total_equals_off_arr
sdis_enhanced$n_total_estimated[mask_off_arr] <- sdis_enhanced$n_total_reported[mask_off_arr]
sdis_enhanced$n_total_estimated_comment[mask_off_arr] <- "Used reported total (matches offenders + arrestees)"

# Rule 2: States where n_total == n_offenders + n_arrestees + n_forensic
mask_all <- sdis_enhanced$total_equals_all
sdis_enhanced$n_total_estimated[mask_all] <- sdis_enhanced$n_total_reported[mask_all] - sdis_enhanced$n_forensic[mask_all]
sdis_enhanced$n_total_estimated_comment[mask_all] <- "Subtracted forensic from reported total"

# Rule 3: States where n_total == n_offenders
mask_off_only <- sdis_enhanced$total_equals_offenders
sdis_enhanced$n_total_estimated[mask_off_only] <- sdis_enhanced$n_total_reported[mask_off_only]
sdis_enhanced$n_total_estimated_comment[mask_off_only] <- "Used reported total (matches offenders only)"

# Rule 4: For remaining states with n_total
mask_total_only <- !is.na(sdis_enhanced$n_total_reported) & 
  is.na(sdis_enhanced$n_total_estimated) &
  (is.na(sdis_enhanced$n_arrestees) | sdis_enhanced$n_arrestees == 0) &
  (is.na(sdis_enhanced$n_offenders) | sdis_enhanced$n_offenders == 0) &
  (is.na(sdis_enhanced$n_forensic) | sdis_enhanced$n_forensic == 0)

sdis_enhanced$n_total_estimated[mask_total_only] <- sdis_enhanced$n_total_reported[mask_total_only]
sdis_enhanced$n_total_estimated_comment[mask_total_only] <- "Total only reported (no breakdown available)"

# Special case: States with n_total and n_forensic only
mask_total_forensic_only <- !is.na(sdis_enhanced$n_total_reported) & 
  is.na(sdis_enhanced$n_total_estimated) &
  (is.na(sdis_enhanced$n_arrestees) | sdis_enhanced$n_arrestees == 0) &
  (is.na(sdis_enhanced$n_offenders) | sdis_enhanced$n_offenders == 0) &
  !is.na(sdis_enhanced$n_forensic) & sdis_enhanced$n_forensic > 0

sdis_enhanced$n_total_estimated[mask_total_forensic_only] <- sdis_enhanced$n_total_reported[mask_total_forensic_only]
sdis_enhanced$n_total_estimated_comment[mask_total_forensic_only] <- "Total only reported (forensic reported separately)"

# States with total and some components but unclear calculation
mask_has_total_unclear <- !is.na(sdis_enhanced$n_total_reported) & is.na(sdis_enhanced$n_total_estimated)
sdis_enhanced$n_total_estimated[mask_has_total_unclear] <- sdis_enhanced$n_total_reported[mask_has_total_unclear]
sdis_enhanced$n_total_estimated_comment[mask_has_total_unclear] <- "Total with discrepancy (calculation unclear)"

# For states without any total but with offenders and arrestees
mask_no_total <- is.na(sdis_enhanced$n_total_reported) & 
  !is.na(sdis_enhanced$n_offenders) & sdis_enhanced$n_offenders > 0 &
  !is.na(sdis_enhanced$n_arrestees) & sdis_enhanced$n_arrestees > 0

sdis_enhanced$n_total_estimated[mask_no_total] <- sdis_enhanced$n_offenders[mask_no_total] + sdis_enhanced$n_arrestees[mask_no_total]
sdis_enhanced$n_total_estimated_comment[mask_no_total] <- "Calculated from offenders + arrestees (no total reported)"

# For states without total but with only offenders > 0
mask_no_total_off_only <- is.na(sdis_enhanced$n_total_reported) & 
  !is.na(sdis_enhanced$n_offenders) & sdis_enhanced$n_offenders > 0 &
  (is.na(sdis_enhanced$n_arrestees) | sdis_enhanced$n_arrestees == 0)

sdis_enhanced$n_total_estimated[mask_no_total_off_only] <- sdis_enhanced$n_offenders[mask_no_total_off_only]
sdis_enhanced$n_total_estimated_comment[mask_no_total_off_only] <- "Used offenders count (no total reported, no arrestee data)"

# Create enhanced data availability matrix with better visualization
availability_matrix <- sdis_enhanced %>%
  transmute(
    State = state,
    Arrestees = ifelse(!is.na(n_arrestees) & n_arrestees > 0, "✓", "!"),
    Offenders = ifelse(!is.na(n_offenders) & n_offenders > 0, "✓", "!"),
    Forensic = ifelse(!is.na(n_forensic) & n_forensic > 0, "✓", "!"),
    `Total Reported` = ifelse(!is.na(n_total_reported), "✓", "!"),
    `Total Method` = case_when(
      total_method == "Offenders only" ~ "O",
      total_method == "Offenders + Arrestees" ~ "O+A",
      total_method == "All components" ~ "All",
      TRUE ~ "?"
    )
  )

# Convert to long format for ggplot
availability_long <- availability_matrix %>%
  pivot_longer(cols = -State, names_to = "Field", values_to = "Value") %>%
  mutate(
    State = factor(State, levels = sort(unique(State), decreasing = TRUE)),
    Field = factor(Field, levels = c("Arrestees", "Offenders", "Forensic", "Total Reported", "Total Method")),
    # Create numeric values for coloring
    NumericValue = case_when(
      Value == "✓" ~ 1,
      Value == "!" ~ 0,
      Value == "O" ~ 2,
      Value == "O+A" ~ 3,
      Value == "All" ~ 4,
      Value == "?" ~ 0,
      TRUE ~ 0
    ),
    # Create display labels
    DisplayLabel = case_when(
      Field == "Total Method" & Value == "?" ~ "?",
      Field == "Total Method" ~ Value,
      TRUE ~ Value
    )
  )

# Summary of how n_total_estimated was calculated
cat("n_total_estimated Column Calculation Summary:\n")
cat("=", strrep("=", 48), "\n", sep = "")

# Group states by how their n_total_estimated was determined
estimation_groups <- sdis_enhanced %>%
  group_by(n_total_estimated_comment) %>%
  summarise(
    states = list(state),
    count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(count))

# Print summary in organized sections
cat("\n1. States with n_total reported and matching patterns:\n")
matching_patterns <- estimation_groups %>%
  filter(str_detect(n_total_estimated_comment, "matches|Subtracted"))
for (i in 1:nrow(matching_patterns)) {
  group <- matching_patterns[i, ]
  cat(sprintf("   • %s: %d states \n(%s)\n", 
              group$n_total_estimated_comment, 
              group$count,
              paste(unlist(group$states), collapse = ", ")))
}

cat("\n2. States with only n_total reported (no breakdown):\n")
total_only <- estimation_groups %>%
  filter(str_detect(n_total_estimated_comment, "Total only reported"))
for (i in 1:nrow(total_only)) {
  group <- total_only[i, ]
  cat(sprintf("   • %s: %d states \n(%s)\n", 
              group$n_total_estimated_comment, 
              group$count,
              paste(unlist(group$states), collapse = ", ")))
}

cat("\n3. States without n_total reported (calculated):\n")
calculated <- estimation_groups %>%
  filter(str_detect(n_total_estimated_comment, "Calculated|Used offenders count"))
for (i in 1:nrow(calculated)) {
  group <- calculated[i, ]
  cat(sprintf("   • %s: %d states (%s)\n", 
              group$n_total_estimated_comment, 
              group$count,
              paste(unlist(group$states), collapse = ", ")))
}

cat("\n4. States with unclear calculation patterns:\n")
unclear <- estimation_groups %>%
  filter(str_detect(n_total_estimated_comment, "discrepancy|unclear"))
for (i in 1:nrow(unclear)) {
  group <- unclear[i, ]
  cat(sprintf("   • %s: %d states (%s)\n", 
              group$n_total_estimated_comment, 
              group$count,
              paste(unlist(group$states), collapse = ", ")))
  
  # Show details for unclear states
  unclear_states <- sdis_enhanced %>% 
    filter(state %in% unlist(group$states)) %>%
    select(state, n_total_reported, n_offenders, n_arrestees, n_forensic)
  
  for (j in 1:nrow(unclear_states)) {
    s <- unclear_states[j, ]
    cat(sprintf("      - %s: Total=%s, Offenders=%s, Arrestees=%s, Forensic=%s\n",
                s$state, 
                format(s$n_total_reported %||% 0, big.mark = ","),
                format(s$n_offenders %||% 0, big.mark = ","),
                format(s$n_arrestees %||% 0, big.mark = ","),
                format(s$n_forensic %||% 0, big.mark = ",")))
  }
}

# Create enhanced heatmap with legend at top and x-labels on both top and bottom
p <- ggplot(availability_long, aes(x = Field, y = State, fill = factor(NumericValue))) +
  geom_tile(color = "white", linewidth = 0.8, width = 0.9, height = 0.9) +
  geom_text(aes(label = DisplayLabel, color = factor(NumericValue)), 
            size = 3.5, fontface = "bold", vjust = 0.8) +
  scale_fill_manual(
    name = "Legend",
    values = c(
      "0" = "#FF6B6B",      # Red/Orange for missing/unknown
      "1" = "#2E86AB",      # Blue for available
      "2" = "#4ECDC4",      # Teal for Offenders only
      "3" = "#45B7D1",      # Light blue for Offenders + Arrestees
      "4" = "#1A535C"       # Dark teal for All components
    ),
    labels = c(
      "0" = "Missing/Unknown data",
      "1" = "Data available",
      "2" = "Total = Offenders only",
      "3" = "Total = Offenders + Arrestees",
      "4" = "Total = All components"
    )
  ) +
  scale_color_manual(values = c(
    "0" = "white",        # White text on red
    "1" = "white",        # White text on blue
    "2" = "white",        # White text on teal
    "3" = "white",        # White text on light blue
    "4" = "white"         # White text on dark teal
  )) +
  labs(
    title = " ",
    x = "", 
    y = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x.top = element_text(angle = 315, hjust = 0.5, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40", size = 9),
    panel.grid = element_blank(),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9),
    axis.ticks.x = element_line(),
    axis.ticks.x.top = element_line()
  ) +
  guides(color = "none") +
  coord_fixed() +
  scale_x_discrete(position = "top")

# Update the main dataframe for subsequent analyses
sdis_data <- sdis_enhanced

# Helper function for null coalescing
`%||%` <- function(a, b) if (!is.null(a) && !is.na(a)) a else b
```

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 30

print(p)
```

```{r}
#| label: summary-visualization
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show summary visualization code"
#| fig-width: 12
#| fig-height: 10

# Create summary data
summary_data <- sdis_data %>%
  mutate(
    data_category = case_when(
      str_detect(n_total_estimated_comment, "matches offenders only") ~ "Matches Offenders Only",
      str_detect(n_total_estimated_comment, "matches offenders \\+ arrestees") ~ "Matches Offenders + Arrestees",
      str_detect(n_total_estimated_comment, "Total only reported \\(no breakdown\\)") ~ "Total Only (No Breakdown)",
      str_detect(n_total_estimated_comment, "Total only reported \\(forensic") ~ "Total Only (Forensic Separate)",
      str_detect(n_total_estimated_comment, "Calculated from offenders \\+ arrestees") ~ "Calculated (Offenders + Arrestees)",
      str_detect(n_total_estimated_comment, "Used offenders count") ~ "Calculated (Offenders Only)",
      str_detect(n_total_estimated_comment, "discrepancy|unclear") ~ "Unclear Calculation",
      TRUE ~ "Other"
    ),
    has_arrestees = !is.na(n_arrestees) & n_arrestees > 0,
    has_offenders = !is.na(n_offenders) & n_offenders > 0,
    has_forensic = !is.na(n_forensic) & n_forensic > 0,
    has_total_reported = !is.na(n_total_reported)
  )

# Create a summary table by category
category_summary <- summary_data %>%
  group_by(data_category) %>%
  summarise(
    count = n(),
    states = paste(sort(state), collapse = ", "),
    avg_total = mean(n_total_estimated, na.rm = TRUE),
    median_total = median(n_total_estimated, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(count))

# Print summary statistics
cat("SUMMARY STATISTICS\n")
cat("=================\n\n")

cat("Calculation Methods Distribution:\n")
for (i in 1:nrow(category_summary)) {
  cat(sprintf("\n%s: %d states (%s)\n", 
              category_summary$data_category[i], 
              category_summary$count[i],
              category_summary$states[i]))
}

cat("\nOverall Data Availability:\n")
cat(sprintf("States with arrestee data: %d\n", sum(summary_data$has_arrestees)))
cat(sprintf("States with offender data: %d\n", sum(summary_data$has_offenders)))
cat(sprintf("States with forensic data: %d\n", sum(summary_data$has_forensic)))
cat(sprintf("States with total reported: %d\n", sum(summary_data$has_total_reported)))

```

## Stacked Bar Chart: DNA Profile Composition by State

This visualization shows the composition of DNA profiles across states, with separate colors for offender, arrestee, and forensic profiles. States without detailed breakdowns show total counts (reported or calculated).

```{r}
#| label: stacked-bar-chart
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show stacked bar chart code"
#| fig-width: 14
#| fig-height: 10

# Prepare data for stacked bar chart
bar_data <- sdis_data %>%
  mutate(
    # Determine if state has detailed breakdown
    has_breakdown = (!is.na(n_offenders) & n_offenders > 0) | 
                    (!is.na(n_arrestees) & n_arrestees > 0) | 
                    (!is.na(n_forensic) & n_forensic > 0),
    
    # For states with breakdown, use actual values
    offenders_display = ifelse(has_breakdown & !is.na(n_offenders), n_offenders, 0),
    arrestees_display = ifelse(has_breakdown & !is.na(n_arrestees), n_arrestees, 0),
    forensic_display = ifelse(has_breakdown & !is.na(n_forensic), n_forensic, 0),
    
    # For states without breakdown, use total
    total_reported_display = ifelse(!has_breakdown & !is.na(n_total_reported), 
                                    n_total_reported, 0),
    total_calculated_display = ifelse(!has_breakdown & is.na(n_total_reported) & 
                                      !is.na(n_total_estimated), 
                                      n_total_estimated, 0)
  ) %>%
  filter(!is.na(n_total_estimated) | has_breakdown) %>%
  select(state, offenders_display, arrestees_display, forensic_display, 
         total_reported_display, total_calculated_display) %>%
  pivot_longer(cols = -state, names_to = "profile_type", values_to = "count") %>%
  filter(count > 0) %>%
  mutate(
    profile_type = factor(profile_type, 
                         levels = c("offenders_display", "arrestees_display", 
                                   "forensic_display", "total_reported_display", 
                                   "total_calculated_display"),
                         labels = c("Offender Profiles", "Arrestee Profiles", 
                                   "Forensic Profiles", "Total (Reported)", 
                                   "Total (Calculated)"))
  )

# Calculate total for ordering states
state_totals <- bar_data %>%
  group_by(state) %>%
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total))

# Reorder states by total count
bar_data <- bar_data %>%
  mutate(state = factor(state, levels = state_totals$state))

# Create custom label function for y-axis
label_k_m <- function(x) {
  ifelse(x >= 1e6, 
         paste0(round(x / 1e6, 1), "M"),
         ifelse(x >= 1e3,
                paste0(round(x / 1e3, 0), "k"),
                as.character(x)))
}

# Create stacked bar chart
# Create stacked bar chart
ggplot(bar_data, aes(x = state, y = count, fill = profile_type)) +
  geom_bar(stat = "identity", width = 0.8, color = "white", linewidth = 0.2) +
  scale_fill_manual(
    name = "Profile Type",
    values = c(
      "Offender Profiles" = "#2E86AB",
      "Arrestee Profiles" = "#F9A03F",
      "Forensic Profiles" = "#2ca02c",
      "Total (Reported)" = "#CCCCCC",
      "Total (Calculated)" = "#808080"
    )
  ) +
  scale_y_continuous(
    labels = label_k_m,
    breaks = seq(0, max(bar_data %>% group_by(state) %>% summarise(total = sum(count)) %>% pull(total), na.rm = TRUE), 
                 by = 250000)
  ) +
  labs(
    title = " ",
    x = NULL,
    y = "Number of Profiles"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0, size = 15, vjust = 0, margin = margin(t = 0)),
    axis.text.y = element_text(size = 15, margin = margin(t = 0)),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 18),
    legend.position = "top",
    legend.spacing.x = unit(1, "cm"),
    legend.spacing.y = unit(0.2, "cm"),
    legend.box.spacing = unit(0.2, "cm"),
    legend.text = element_text(size = 15),
    legend.title = element_text(face = "bold", size = 18),
    axis.title.x = element_text(size = 18, margin = margin(t = 10)),
    axis.title.y = element_text(size = 18, margin = margin(r = 10)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE))
```

## Export Enhanced Dataset

Export the enhanced dataset with the new n_total_estimated values and documentation.

```{r}
#| label: export-enhanced-data
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show exportation code"


# Prepare final dataset with key columns in logical order
final_columns <- c(
  'state', 
  'n_total_estimated',
  'n_total_reported',
  'n_total_estimated_comment',
  'total_method',
  'n_arrestees', 
  'n_offenders', 
  'n_forensic',
  'arrestee_collection',
  'fam_search',
  'collection_statute'
)

# Select columns that exist in the dataset
available_columns <- intersect(final_columns, names(sdis_enhanced))
sdis_final <- sdis_enhanced %>% select(all_of(available_columns))

# Export to CSV
output_relative <- file.path("data", "sdis", "final", "SDIS_cross_section.csv")
output_path <- here(output_relative)
write_csv(sdis_final, output_path)
cat(paste("Exported enhanced SDIS dataset to:", output_relative, "\n"))

# Display final summary statistics
cat("\n\nFinal Summary Statistics:\n")
cat("=", strrep("=", 48), "\n", sep = "")
cat(paste("Total states in dataset:", nrow(sdis_final), "\n"))
cat(paste("States with n_total_estimated:", sum(!is.na(sdis_final$n_total_estimated)), "\n"))
cat(paste("States with n_total_reported:", sum(!is.na(sdis_final$n_total_reported)), "\n"))
cat(paste("Total profiles (estimated):", format(sum(sdis_final$n_total_estimated, na.rm = TRUE), big.mark = ","), "\n"))

# Prepare the data for the interactive table
summary_table <- sdis_data %>%
  select(state, n_total_estimated, n_total_reported, n_offenders, n_arrestees, 
         n_forensic, n_total_estimated_comment) %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), NA, format(., big.mark = ",", scientific = FALSE))))

# Create interactive table
datatable(
  summary_table,
  extensions = c('Buttons', 'ColReorder', 'Scroller'),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'colvis'),
    scrollX = TRUE,
    scrollY = "600px",
    scroller = TRUE,
    pageLength = 20,
    columnDefs = list(
      list(className = 'dt-right', targets = 1:5),  # Right-align numeric columns
      list(className = 'dt-left', targets = c(0, 6))  # Left-align state and comment columns
    )
  ),
  rownames = FALSE,
  filter = 'top',
  caption = "Enhanced Data with Estimated Totals"
)

# Create final frozen version (v1.0)
frozen_dir_relative <- file.path("data", "v1.0")
frozen_dir <- here(frozen_dir_relative)
dir.create(frozen_dir, recursive = TRUE, showWarnings = FALSE)

frozen_path_relative <- file.path(frozen_dir_relative, "SDIS_cross_section.csv")
frozen_path <- here(frozen_path_relative)
write_csv(sdis_final, frozen_path)
cat(paste("✓ Created frozen version 1.0 at:", frozen_path_relative, "\n"))

```

# Comparison between NDIS and SDIS datasets

```{r}
#| label: NDIS-SDIS-data
#| echo: true
#| code-fold: true
#| results: hold
#| code-summary: "Show NDIS and SDIS comparison code"
library(tidyverse)
library(here)
library(scales)
library(patchwork)

# Read frozen datasets
ndis <- read_csv(here("data", "v1.0", "NDIS_time_series.csv"))
sdis <- read_csv(here("data", "v1.0", "SDIS_cross_section.csv"))

# Check what date column exists in NDIS data
date_col <- if("year_month" %in% names(ndis)) {
  "year_month"
} else if("date" %in% names(ndis)) {
  "date"
} else if("snapshot_date" %in% names(ndis)) {
  "snapshot_date"
} else {
  # If no date column, we can't filter by latest
  NULL
}

# Extract latest NDIS data for each jurisdiction
if (!is.null(date_col)) {
  ndis_latest <- ndis %>%
    group_by(jurisdiction) %>%
    filter(.data[[date_col]] == max(.data[[date_col]])) %>%
    ungroup() %>%
    select(jurisdiction, ndis_profiles = total_profiles)
} else {
  # If no date column, just take the data as-is (assuming it's already latest)
  ndis_latest <- ndis %>%
    select(jurisdiction, ndis_profiles = total_profiles) %>%
    distinct(jurisdiction, .keep_all = TRUE)
}

# Prepare SDIS data
sdis_summary <- sdis %>%
  mutate(sdis_profiles = if_else(is.na(n_total_reported), n_total_estimated, n_total_reported)) %>%
  select(jurisdiction = state, sdis_profiles)

# Combine datasets
combined <- full_join(ndis_latest, sdis_summary, by = "jurisdiction") %>%
  mutate(
    ndis_profiles = replace_na(ndis_profiles, 0),
    sdis_profiles = replace_na(sdis_profiles, NA)
  )

plot_data <- combined %>%
  pivot_longer(cols = c(ndis_profiles, sdis_profiles),
               names_to = "source", values_to = "profiles") %>%
  mutate(
    profiles = replace_na(profiles, 0),
    source = factor(source, levels = c("sdis_profiles", "ndis_profiles"), 
                    labels = c("SDIS", "NDIS"))
  )

# Divide into two groups
high_states <- plot_data %>%
  group_by(jurisdiction) %>%
  summarise(total = sum(profiles, na.rm = TRUE)) %>%
  filter(total > 400000) %>%
  pull(jurisdiction) %>%
  sort()

low_states <- plot_data %>%
  filter(!jurisdiction %in% high_states) %>%
  pull(jurisdiction) %>%
  unique() %>%
  sort()

# Plot function with publication-quality styling
plot_profiles <- function(data, states, title) {
  data <- data %>%
    mutate(jurisdiction = factor(jurisdiction, levels = rev(states)))
  
  ggplot(data %>% filter(jurisdiction %in% states),
         aes(y = jurisdiction, x = profiles, fill = source)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6, 
             color = "black", linewidth = 0.3) +
    scale_fill_manual(
      values = c("NDIS" = "#1f77b4", "SDIS" = "#ff7f0e"),
      labels = c("NDIS", "SDIS")
    ) +
    scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
    labs(title = title, y = "Jurisdiction", x = "Total Profiles", fill = "Source") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.y = element_text(size = 12, color = "black"),
      axis.text.x = element_text(size = 12, color = "black"),
      axis.title = element_text(size = 13, face = "bold"),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.position = "top",
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      panel.grid.major.x = element_line(color = "gray90", linewidth = 0.3),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
      axis.line = element_line(color = "black", linewidth = 0.3),
      axis.ticks = element_line(color = "black", linewidth = 0.3),
      axis.ticks.length = unit(0.15, "cm"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

# Plot for high profile states (alphabetical order, reversed for y-axis)
p_high <- plot_profiles(plot_data, high_states, "States with >400k Total Profiles")

# Plot for low profile states (alphabetical order, reversed for y-axis)
p_low <- plot_profiles(plot_data, low_states, "States with ≤400k Total Profiles")
```

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 12

# Show plots side by side
p_high | p_low
```
