---
title: "Annual DNA collection from Murphy & Tong Study"
author: "Tina Lasisi & João P. Donadio"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| echo: false
#| message: false
#| warning: false

# List of required packages
required_packages <- c(
  "tidyverse",   # Data manipulation and visualization
  "knitr",       # Dynamic report generation
  "plotly",      # Interactive plots
  "DT",          # Interactive tables
  "dplyr",       # Data manipulation
  "stringr",     # String manipulation
  "purrr",       # Functional programming
  "here",        # File path management
  "flextable",   # Table formatting
  "kableExtra"   # Table formatting for PDF
)

# Function to install missing packages
install_missing <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      message(paste("Installing missing package:", pkg))
      install.packages(pkg, dependencies = TRUE)
    }
  }
}

# Install any missing packages
install_missing(required_packages)

# Load all packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(DT)
  library(plotly)
  library(dplyr)
  library(stringr)
  library(purrr)
  library(here)
  library(flextable)
  library(kableExtra)
})

# Verify all packages loaded successfully
loaded_packages <- sapply(required_packages, require, character.only = TRUE)

if (all(loaded_packages)) {
  message("📚 All packages loaded successfully!")
} else {
  warning("The following packages failed to load: ", 
          paste(names(loaded_packages)[!loaded_packages], collapse = ", "))
}
```

## Executive Summary

This document provides complete transparency regarding the data sources and methodology used to compile racial disparities in DNA collection across U.S. states. The original data was collected in Summer 2017, with most data points from 2013-2016.

### Key Findings

-   **Consistent racial disparities**: Black populations show the highest DNA collection rates relative to their population percentage in nearly all states

-   **Data limitations**: Many states lack comprehensive conviction data, requiring the use of prison admission proxies

-   **Methodological challenges**: Hispanic/Latino populations often uncounted or miscategorized in state data

## Methodology Overview

### Data Collection Period

-   Primary collection: Summer 2017

-   Data years used: Single year per state (2012-2016, varies by availability)

-   Census baseline: 2010 U.S. Census for demographic comparisons

### General Challenges Encountered

1.   **Conviction Data Scarcity**: Most states do not publicly disclose comprehensive felony conviction data

2.   **Prison Admission Proxy**: Prison admissions used as substitute for conviction data in majority of states

3.   **Racial Classification Inconsistencies**:

    -   Many states only report "Black" and "White" categories

    -   Hispanic/Latino often classified as ethnicity rather than race

    -   "Other" category frequently used without specification

4.   **Arrest Data Gaps**: Racial makeup of arrests often unavailable or incomplete

## National Summary Table

### Methodology: Parsing DNA Collection Data

The national summary data was extracted from a structured text file (`MurphyTong_Racial_Breakdown.txt`) containing three distinct data sections for each state:

1. **DNA Collection Data**: Percentage and absolute counts of DNA profiles collected by race (e.g., "46% B (18,253)")
2. **Population Demographics**: Percentage of state population by race from 2010 Census

**Processing Steps**:

- **State Identification**: The parser identifies state entries using standard two-letter abbreviations (AL, AK, AZ, etc.)
- **Section Segmentation**: For each state, the text is divided into three sections based on the pattern of "% B" (Black percentage) occurrences
- **Data Extraction**: Regular expressions extract both percentages and counts from the first section, and percentages only from the demographic and rate sections
- **Pattern Matching**: The code uses regex patterns like `([0-9.]+)%\s*B\s*\(([0-9,]+)\)` to capture both percentage (e.g., 46%) and count (e.g., 18,253) for DNA collection data
- **Race Categories**: Data is extracted for five racial categories: Black (B), Hispanic (H), Asian (A), Native American (N), and White (W)

This automated parsing ensures reproducibility and allows for systematic extraction of data from the original Murphy & Tong study format.

```{r}
#| label: tbl-national-summary
#| tbl-cap: "Racial Breakdown of Annual DNA Collection for Each State"
#| echo: true
#| code-fold: true
#| code-summary: "Show summary code"
#| results: hold

# Function to parse DNA collection data from text file
parse_dna_data <- function(file_path) {
  
  # Read the entire file
  text_data <- readLines(file_path, warn = FALSE)
  
  # Remove empty lines
  text_data <- text_data[text_data != ""]
  
  # Initialize list to store parsed data
  parsed_data <- list()
  
  # State abbreviations (for reference)
  state_abbrevs <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
                     "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                     "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
                     "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
                     "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "DC")
  
  # Function to extract percentage and count from patterns like "46% B (18,253)"
  extract_race_data <- function(text, race_letter) {
    pattern <- paste0("([0-9.]+)%\\s*", race_letter, "\\s*\\(([0-9,]+)\\)")
    matches <- str_match(text, pattern)
    if (!is.na(matches[1])) {
      pct <- as.numeric(matches[2])
      count <- as.numeric(gsub(",", "", matches[3]))
      return(list(pct = pct, count = count))
    }
    return(list(pct = NA, count = NA))
  }
  
  # Function to extract just percentage (for demographics and collection rates)
  extract_percentage <- function(text, race_letter) {
    pattern <- paste0("([0-9.]+)%\\s*", race_letter)
    matches <- str_match(text, pattern)
    if (!is.na(matches[1])) {
      return(as.numeric(matches[2]))
    }
    return(NA)
  }
  
  # Process each line
  i <- 1
  while (i <= length(text_data)) {
    line <- text_data[i]
    
    # Check if line is a state abbreviation
    if (line %in% state_abbrevs) {
      state <- line
      
      # Initialize data structure for this state
      state_data <- list(
        State = state,
        Black_DNA_Pct = NA, Black_DNA_N = NA,
        Hispanic_DNA_Pct = NA, Hispanic_DNA_N = NA,
        Asian_DNA_Pct = NA, Asian_DNA_N = NA,
        Native_American_DNA_Pct = NA, Native_American_DNA_N = NA,
        White_DNA_Pct = NA, White_DNA_N = NA,
        Black_Pop_Pct = NA, Hispanic_Pop_Pct = NA, Asian_Pop_Pct = NA,
        Native_American_Pop_Pct = NA, White_Pop_Pct = NA,
        Black_Collection_Rate = NA, Hispanic_Collection_Rate = NA,
        Asian_Collection_Rate = NA, Native_American_Collection_Rate = NA,
        White_Collection_Rate = NA
      )
      
      # Collect all lines for this state until we hit the next state or end of file
      state_lines <- c()
      i <- i + 1
      while (i <= length(text_data) && !(text_data[i] %in% state_abbrevs)) {
        state_lines <- c(state_lines, text_data[i])
        i <- i + 1
      }
      
      # Combine all lines for this state into one text block
      state_text <- paste(state_lines, collapse = " ")
      
      # Split into sections based on the pattern of % B occurrences
      # We need to identify the three sections: DNA Collection, Demographics, Collection Rates
      
      # Find all "% B" patterns to help identify sections
      b_patterns <- str_locate_all(state_text, "[0-9.]+%\\s*B")[[1]]
      
      if (nrow(b_patterns) >= 1) {
        # First section: DNA Collection (has counts in parentheses)
        dna_section_start <- 1
        dna_section_end <- if(nrow(b_patterns) >= 2) b_patterns[2,1] - 1 else nchar(state_text)
        dna_section <- substr(state_text, dna_section_start, dna_section_end)
        
        # Extract DNA collection data (with counts)
        black_dna <- extract_race_data(dna_section, "B")
        hispanic_dna <- extract_race_data(dna_section, "H")
        asian_dna <- extract_race_data(dna_section, "A")
        native_dna <- extract_race_data(dna_section, "N")
        white_dna <- extract_race_data(dna_section, "W")
        
        state_data$Black_DNA_Pct <- black_dna$pct
        state_data$Black_DNA_N <- black_dna$count
        state_data$Hispanic_DNA_Pct <- hispanic_dna$pct
        state_data$Hispanic_DNA_N <- hispanic_dna$count
        state_data$Asian_DNA_Pct <- asian_dna$pct
        state_data$Asian_DNA_N <- asian_dna$count
        state_data$Native_American_DNA_Pct <- native_dna$pct
        state_data$Native_American_DNA_N <- native_dna$count
        state_data$White_DNA_Pct <- white_dna$pct
        state_data$White_DNA_N <- white_dna$count
      }
      
      if (nrow(b_patterns) >= 2) {
        # Second section: Demographics (percentages only, no parentheses)
        demo_section_start <- b_patterns[2,1]
        demo_section_end <- if(nrow(b_patterns) >= 3) b_patterns[3,1] - 1 else nchar(state_text)
        demo_section <- substr(state_text, demo_section_start, demo_section_end)
        
        # Extract demographic percentages
        state_data$Black_Pop_Pct <- extract_percentage(demo_section, "B")
        state_data$Hispanic_Pop_Pct <- extract_percentage(demo_section, "H")
        state_data$Asian_Pop_Pct <- extract_percentage(demo_section, "A")
        state_data$Native_American_Pop_Pct <- extract_percentage(demo_section, "N")
        state_data$White_Pop_Pct <- extract_percentage(demo_section, "W")
      }
      
      if (nrow(b_patterns) >= 3) {
        # Third section: Collection Rates (percentages only, no parentheses)
        rate_section_start <- b_patterns[3,1]
        rate_section <- substr(state_text, rate_section_start, nchar(state_text))
        
        # Extract collection rate percentages
        state_data$Black_Collection_Rate <- extract_percentage(rate_section, "B")
        state_data$Hispanic_Collection_Rate <- extract_percentage(rate_section, "H")
        state_data$Asian_Collection_Rate <- extract_percentage(rate_section, "A")
        state_data$Native_American_Collection_Rate <- extract_percentage(rate_section, "N")
        state_data$White_Collection_Rate <- extract_percentage(rate_section, "W")
      }
      
      # Add to parsed data
      parsed_data[[length(parsed_data) + 1]] <- state_data
      
      # Don't increment i here since we already did it in the while loop
      i <- i - 1
    }
    
    i <- i + 1
  }
  
  # Convert to data frame
  df <- do.call(rbind, lapply(parsed_data, data.frame))
  
  return(df)
}

racial_data_path <- file.path(here("data", "annual_dna_collection", "intermediate", "MurphyTong_Racial_Breakdown.txt"))

summary_data <- parse_dna_data(racial_data_path)

# Display interactive table
datatable(summary_data, 
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Complete state-by-state breakdown of DNA collection by race")
```

### Disparity Analysis

```{r}
#| label: fig-disparity-rates
#| fig-cap: "DNA Collection Rates by Race Relative to Population Percentage"
#| echo: true
#| code-fold: true
#| code-summary: "Show disparity analysis code"

# Calculate disparity ratios
disparity_data <- summary_data %>%
  filter(!is.na(Black_Collection_Rate) & !is.na(White_Collection_Rate)) %>%
  mutate(Black_White_Ratio = Black_Collection_Rate / White_Collection_Rate) %>%
  arrange(desc(Black_White_Ratio))

# Create visualization
ggplot(disparity_data %>% head(20), aes(x = reorder(State, Black_White_Ratio), y = Black_White_Ratio)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(title = "Top 20 States: Black-White DNA Collection Disparity Ratio",
       subtitle = "Ratio of collection rates (higher = greater disparity)",
       x = "State",
       y = "Black/White Collection Rate Ratio") +
  theme_minimal() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50")

```

## State-by-State Detailed Methodology

### Methodology: Parsing State Methodology Paragraphs

The detailed methodology for each state was extracted from a separate text file (`MurphyTong_States_Paragraphs.txt`) containing narrative descriptions of data collection approaches for all 50 states. This section explains how we systematically parsed this unstructured text into a structured dataset.

**Processing Steps**:

1. **State Detection**: The parser identifies state entries by searching for the 50 U.S. state names as they appear in the text
2. **Section Extraction**: For each state, the parser captures all text from the state name until the next state name appears
3. **Component Parsing**: Within each state's section, the code extracts four key components:
   - **Legal Framework**: The statutory basis for DNA collection in that state
   - **Collection Triggers**: Specific offenses or events that trigger DNA collection
   - **Data Sources**: Types of data used (e.g., conviction records, prison admissions, arrest data)
   - **Source URLs**: Web links to the original data sources
   - **Data Limitations**: Known gaps, proxies, or methodological caveats

4. **Structured Data Creation**: Each data source line is parsed into a type-note pair (e.g., "Prison admissions: Used as proxy for conviction data")

5. **Categorization**: The parser automatically categorizes:
   - **Collection trigger types**: Comprehensive, selective, felony-only, etc.
   - **Data limitation types**: Missing conviction data, ethnicity issues, limited racial categories, etc.
   - **Data source types**: Conviction data, arrest data, prison data, sex crime data, etc.

This systematic extraction allows for consistent comparison across states and identification of common patterns in data collection methodologies and limitations.

```{r}
#| label: sec-state
#| echo: true
#| code-fold: true
#| code-summary: "Show paragraph extraction code"
#| results: hold

# Pre-define all 50 U.S. states
us_states <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", 
               "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", 
               "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", 
               "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", 
               "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
               "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
               "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", 
               "Wisconsin", "Wyoming")

# Read the text file
data_file <- file.path(here("data", "annual_dna_collection", "intermediate", "MurphyTong_States_Paragraphs.txt"))

text_content <- readLines(data_file, warn = FALSE)

# Combine all lines into a single string
full_text <- paste(text_content, collapse = "\n")

# Create a pattern to match state names
state_pattern <- paste0("(?m)^[[:space:]]*(", paste(us_states, collapse = "|"), ")\\b")
state_matches <- str_locate_all(full_text, state_pattern)[[1]]

# Extract state sections
state_sections <- list()
state_names <- character()

if (nrow(state_matches) > 0) {
  for (i in 1:nrow(state_matches)) {
    start_pos <- state_matches[i, "start"]
    if (i < nrow(state_matches)) {
      end_pos <- state_matches[i + 1, "start"] - 1
    } else {
      end_pos <- nchar(full_text)
    }
    
    state_name <- substr(full_text, start_pos, state_matches[i, "end"])
    state_content <- substr(full_text, start_pos, end_pos)
    
    state_names <- c(state_names, state_name)
    state_sections <- c(state_sections, state_content)
  }
}

# Function to extract information from each state section
parse_state_section <- function(section, state_name) {
  if (is.na(state_name)) return(NULL)
  
  # Extract legal framework
  legal_framework <- str_extract(section, "Legal Framework:[^\n]+") %>%
    {ifelse(is.na(.), NA, str_remove(., "Legal Framework:") %>% str_trim())}
  
  # Extract collection triggers
  collection_triggers <- str_extract(section, "Collection Triggers:[^\n]+") %>%
    {ifelse(is.na(.), NA, str_remove(., "Collection Triggers:") %>% str_trim())}
  
  # Extract data sources - capture everything until Source URLs or Data Limitations
  data_sources_text <- str_extract(section, "Data Sources:[\\s\\S]*?(?=Source URLs:|Data Limitations:|$)")
  data_source_df <- tibble(data_source_type = NA_character_, data_source_note = NA_character_)
  
  if (!is.na(data_sources_text)) {
    data_sources_text <- str_remove(data_sources_text, "Data Sources:") %>% str_trim()
    # Split by newlines and clean up
    data_source_lines <- str_split(data_sources_text, "\\n")[[1]] %>% 
      str_trim() %>% 
      discard(~ .x == "" | str_detect(.x, "^Source URLs:|^Data Limitations:"))
    
    if (length(data_source_lines) > 0) {
      data_source_df <- map_df(data_source_lines, function(line) {
        if (str_detect(line, ":")) {
          tibble(
            data_source_type = str_extract(line, "^[^:]+") %>% str_trim(),
            data_source_note = str_remove(line, "^[^:]+:") %>% str_trim()
          )
        } else {
          tibble(
            data_source_type = line,
            data_source_note = NA_character_
          )
        }
      })
    }
  }
  
  # Extract source URLs
  source_urls_text <- str_extract(section, "Source URLs:[\\s\\S]*?(?=Data Limitations:|$)")
  source_urls <- character(0)
  
  if (!is.na(source_urls_text)) {
    source_urls_text <- str_remove(source_urls_text, "Source URLs:") %>% str_trim()
    source_url_lines <- str_split(source_urls_text, "\\n")[[1]] %>% 
      str_trim() %>% 
      discard(~ .x == "" | str_detect(.x, "^Data Limitations:"))
    
    if (length(source_url_lines) > 0) {
      source_urls <- source_url_lines
    }
  }
  
  # Extract data limitations - capture everything until next state or end
  data_limitations_text <- str_extract(section, "Data Limitations:[\\s\\S]*?(?=\\b(A|Ala|Alas|Ari|Arka|Cali|Colo|Conn|Del|Flo|Geo|Haw|Ida|Ill|Ind|Iow|Kan|Ken|Lou|Mai|Mar|Mas|Mic|Min|Mis|Mon|Neb|Nev|New|Nor|Ohi|Okl|Ore|Pen|Rho|Sou|Ten|Tex|Uta|Ver|Vir|Was|Wis|Wyo)\\b|$)")
  data_limitations <- NA_character_
  
  if (!is.na(data_limitations_text)) {
    data_limitations_text <- str_remove(data_limitations_text, "Data Limitations:") %>% str_trim()
    data_limitations_lines <- str_split(data_limitations_text, "\\n")[[1]] %>% 
      str_trim() %>% 
      discard(~ .x == "" | str_detect(.x, "^\\b(A|Ala|Alas|Ari|Arka|Cali|Colo|Conn|Del|Flo|Geo|Haw|Ida|Ill|Ind|Iow|Kan|Ken|Lou|Mai|Mar|Mas|Mic|Min|Mis|Mon|Neb|Nev|New|Nor|Ohi|Okl|Ore|Pen|Rho|Sou|Ten|Tex|Uta|Ver|Vir|Was|Wis|Wyo)\\b"))
    
    if (length(data_limitations_lines) > 0) {
      data_limitations <- paste(data_limitations_lines, collapse = "; ")
    }
  }
  
  # If no data sources were found, create at least one row for the state
  if (nrow(data_source_df) == 0) {
    data_source_df <- tibble(data_source_type = NA_character_, data_source_note = NA_character_)
  }
  
  # Create result dataframe
  result_df <- data_source_df %>%
    mutate(
      state = state_name,
      legal_framework = legal_framework,
      collection_triggers = collection_triggers,
      source_url = ifelse(length(source_urls) > 0, paste(source_urls, collapse = "; "), NA_character_),
      data_limitations = data_limitations,
      .before = everything()
    )
  
  return(result_df)
}

# Parse all state sections
state_data_list <- lapply(seq_along(state_sections), function(i) {
  parse_state_section(state_sections[[i]], state_names[i])
})

# Combine all results into one dataframe
state_data <- bind_rows(state_data_list)

# Clean up the data - remove rows where all data columns are NA
final_df <- state_data %>%
  mutate(across(where(is.character), ~ ifelse(.x == "" | is.na(.x), NA, .x))) %>%
  filter(!(is.na(data_source_type) & is.na(source_url) & is.na(data_limitations)))

# Fill in the missing legal_framework and other methodology for each state
final_df_clean <- final_df %>%
  group_by(state) %>%
  fill(legal_framework, collection_triggers, .direction = "downup") %>%
  ungroup()

# Now let's fill source_url and data_limitations across all rows for each state
final_df_clean <- final_df_clean %>%
  group_by(state) %>%
  mutate(
    source_url = ifelse(all(is.na(source_url)), NA, 
                       paste(na.omit(unique(source_url)), collapse = "; ")),
    data_limitations = ifelse(all(is.na(data_limitations)), NA,
                             paste(na.omit(unique(data_limitations)), collapse = "; "))
  ) %>%
  ungroup()

# Create categorization columns for easier analysis
final_df_clean <- final_df_clean %>%
  mutate(
    # Categorize collection triggers
    collection_trigger_category = case_when(
      str_detect(collection_triggers, "(?i)all felony.*convictions.*arrests.*all felonies") ~ "Comprehensive: All felonies + broad arrests",
      str_detect(collection_triggers, "(?i)all felony.*convictions.*arrests.*certain|specific") ~ "Selective: All felonies + specific arrests",
      str_detect(collection_triggers, "(?i)all felony.*convictions.*arrests") ~ "Broad: All felonies + various arrests",
      str_detect(collection_triggers, "(?i)all felony.*convictions") ~ "Felony convictions only",
      str_detect(collection_triggers, "(?i)felony.*misdemeanor.*convictions") ~ "Mixed: Felony + misdemeanor convictions",
      TRUE ~ "Other/Unspecified"
    ),
    
    # Categorize data limitations
    data_limitation_category = case_when(
      is.na(data_limitations) ~ "No limitations noted",
      str_detect(data_limitations, "(?i)no direct.*conviction data") ~ "Missing conviction data",
      str_detect(data_limitations, "(?i)prison admissions.*proxy") ~ "Prison data as proxy",
      str_detect(data_limitations, "(?i)hispanic|ethnicity") ~ "Ethnicity categorization issues",
      str_detect(data_limitations, "(?i)racial.*limited|black.*white.*only") ~ "Limited racial categories",
      str_detect(data_limitations, "(?i)no.*data.*available|unavailable") ~ "Various data unavailable",
      TRUE ~ "Other limitations"
    ),
    
    # Categorize data source types
    data_source_category = case_when(
      str_detect(data_source_type, "(?i)conviction") ~ "Conviction Data",
      str_detect(data_source_type, "(?i)arrest") ~ "Arrest Data",
      str_detect(data_source_type, "(?i)sex|sexual") ~ "Sex Crime Data",
      str_detect(data_source_type, "(?i)prison|admission|correction") ~ "Prison/Incarceration Data",
      TRUE ~ "Other Data Source"
    )
  )

# Create a summary table for quick overview - ONE ROW PER STATE
methodology_summary <- final_df_clean %>%
  distinct(state, legal_framework, collection_triggers, collection_trigger_category, 
           data_limitations, data_limitation_category, source_url) %>%
  arrange(state) %>%
  mutate(state = str_trim(state))

# Create interactive table
datatable_methodology <- datatable(
  methodology_summary,
  extensions = c('Buttons', 'ColReorder', 'Scroller'),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'colvis'),
    scrollX = TRUE,
    scrollY = "600px",
    scroller = TRUE,
    pageLength = 10,
    columnDefs = list(
      list(className = 'dt-left', targets = 0:(ncol(methodology_summary)-1)),
      list(width = '200px', targets = c(1, 2, 3))
    ),
    autoWidth = TRUE
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe hover'
) %>%
  formatStyle(
    columns = names(methodology_summary),
    fontSize = '12px',
    lineHeight = '90%'
  )

# Show the table
datatable_methodology
```

## Combining Information into Master Dataset

The final dataset combines quantitative DNA collection metrics with qualitative methodology to create a comprehensive one-row-per-state resource. This integration allows researchers to understand both the magnitude of DNA collection and the quality/limitations of the underlying data sources.

**Integration Process**:

1. **Primary Dataset**: The `summary_data` table contains all quantitative metrics:
   - DNA collection counts and percentages by race
   - State population demographics by race
   - Collection rates per 100,000 population by race

2. **Study Methodology Addition**: The `methodology_summary` table provides contextual information:
   - Legal framework for DNA collection
   - Specific collection triggers
   - Data source types and limitations
   - Source URLs for verification

3. **Joining Strategy**: 
   - States are matched using full state names
   - A crosswalk table converts two-letter abbreviations to full names
   - Left join ensures all states from the summary data are retained

4. **Column Selection**: The final dataset includes:
   - **State identifier**: Full state name
   - **Collection metrics**: All DNA collection counts, percentages, and rates by race
   - **Methodology context**: Legal framework, collection triggers, data limitations
   - **Quality flags**: Categorized data limitation types for filtering/analysis

5. **Output Format**: One row per state with 30+ columns covering demographics, collection rates, and methodology

This unified structure enables analyses that account for data quality differences across states when interpreting racial disparities in DNA collection.

```{r}
#| label: combining-data
#| echo: true
#| code-fold: true
#| code-summary: "Show data combination and exportation code"
#| results: hold

# Create state name crosswalk for joining
state_crosswalk <- tibble(
  State = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", 
            "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
            "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", 
            "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", 
            "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "DC"),
  state_full = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", 
                 "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", 
                 "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", 
                 "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
                 "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", 
                 "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", 
                 "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", 
                 "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", 
                 "Utah", "Vermont", "Virginia", "Washington", "West Virginia", 
                 "Wisconsin", "Wyoming", "District of Columbia")
)

# Join summary data with methodology
annual_dna_combined <- summary_data %>%
  left_join(state_crosswalk, by = "State") %>%
  left_join(methodology_summary, by = c("state_full" = "state")) %>%
  select(
    # State identifier
    state = state_full,
    state_abbrev = State,
    
    # DNA Collection metrics
    Black_DNA_Pct, Black_DNA_N,
    Hispanic_DNA_Pct, Hispanic_DNA_N,
    Asian_DNA_Pct, Asian_DNA_N,
    Native_American_DNA_Pct, Native_American_DNA_N,
    White_DNA_Pct, White_DNA_N,
    
    # Population demographics
    Black_Pop_Pct, Hispanic_Pop_Pct, Asian_Pop_Pct,
    Native_American_Pop_Pct, White_Pop_Pct,
    
    # Collection rates
    Black_Collection_Rate, Hispanic_Collection_Rate,
    Asian_Collection_Rate, Native_American_Collection_Rate,
    White_Collection_Rate,
    
    # Methodology
    legal_framework,
    collection_triggers,
    collection_trigger_category,
    data_limitations,
    data_limitation_category,
    source_url
  )

# Calculate total DNA profiles
annual_dna_combined <- annual_dna_combined %>%
  mutate(across(c(Black_DNA_N, Hispanic_DNA_N, Asian_DNA_N, Native_American_DNA_N, White_DNA_N), ~ replace_na(., 0)),
  Total_DNA_Profiles = Black_DNA_N + Hispanic_DNA_N + Asian_DNA_N + 
                         Native_American_DNA_N + White_DNA_N
  ) %>%
  relocate(Total_DNA_Profiles, .after = White_DNA_N)

# Show interactive preview
datatable(
  annual_dna_combined,
  extensions = c('Buttons', 'ColReorder', 'Scroller'),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'colvis'),
    scrollX = TRUE,
    scrollY = "400px",
    scroller = TRUE,
    pageLength = 10,
    columnDefs = list(
      list(className = 'dt-left', targets = 0:(ncol(annual_dna_combined)-1))
    ),
    autoWidth = TRUE
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe hover',
  caption = "Complete Annual DNA Collection Dataset - One Row Per State"
) %>%
  formatStyle(
    columns = names(annual_dna_combined),
    fontSize = '11px',
    lineHeight = '90%'
  )
```

## Data Export and Versioning

```{r}
#| label: export-data
#| echo: true
#| code-fold: true
#| code-summary: "Show export code"
#| results: hold

# Create output directory structure
intermediate_dir <- here("data", "annual_dna_collection", "intermediate")
dir.create(intermediate_dir, recursive = TRUE, showWarnings = FALSE)

final_dir <- here("data", "annual_dna_collection", "final")
dir.create(final_dir, recursive = TRUE, showWarnings = FALSE)

# Export final combined dataset
output_path <- file.path(final_dir, "Annual_DNA_Collection.csv")
write_csv(annual_dna_combined, output_path)
cat(paste("✓ Exported Annual DNA Collection dataset to:", output_path, "\n"))

# Also export the methodology summary separately for reference
methodology_output_path <- file.path(intermediate_dir, "State_Methodology.csv")
write_csv(methodology_summary, methodology_output_path)
cat(paste("✓ Exported study methodology summary to:", methodology_output_path, "\n"))

cat("Use analysis/version_freeze.qmd to create processed data releases when needed.\n")
```

## Summary and Key Findings

### Dataset Completeness

```{r}
#| label: data-completeness-summary
#| echo: true
#| code-fold: true
#| code-summary: "Show completeness analysis"

# Analyze data completeness by category
completeness_summary <- annual_dna_combined %>%
  summarise(
    States_with_Black_Data = sum(!is.na(Black_Collection_Rate)),
    States_with_Hispanic_Data = sum(!is.na(Hispanic_Collection_Rate)),
    States_with_Asian_Data = sum(!is.na(Asian_Collection_Rate)),
    States_with_Native_American_Data = sum(!is.na(Native_American_Collection_Rate)),
    States_with_White_Data = sum(!is.na(White_Collection_Rate)),
    States_with_Legal_Framework = sum(!is.na(legal_framework))
  )

kable(t(completeness_summary), col.names = "Count", 
      caption = "Data Completeness Across States") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```
